---
title: "Analyzing MLB Pitch Trends"
format: html
editor: visual
---

Run 'quarto render 'Analyzing MLB Pitch Trends.qmd' --output-dir docs' to render docs

In this project we will analyze MLB pitch data from 2015-2018 to view trends in pitching performance using statistical analysis, predictive modeling, and classification techniques. Our goal is to develop models that predict variables like pitch speed while also implementing classification models to determine pitch type.This project will involve data cleaning, visualization, and model evaluation.

We will be using a dataset published on Kaggle titled “MLB Pitch Data (2015-2018),” which contains 8 different tables with information on pitches, at-bats, games, and players. Our focus will primarily be on the “pitches” table, which includes 2,867,154 pitch observations with 40 different attributes describing various aspects of each pitch. While most of the features are continuous variables measuring aspects like speed and rotation, categorical features such as pitch type provide opportunities for classification modeling. The dataset contains noise, including outlier pitches, missing values, and irrelevant features, which will require preprocessing before analysis. By leveraging this data, we aim to build predictive models to analyze pitch characteristics and gain deeper insights into pitching trends.

## Load the CSV

First install any packages we will be using

```{r}
install.packages("corrplot")
install.packages("tidyverse")
```

```{r}
library(tidyverse)

data_file <- "pitches.csv"

pitch_df <- read_csv(data_file)
```

Print the size of each table

```{r}
dim(pitch_df)
```

## Looking into the Pitch Table

Take a look at the features

```{r}
str(pitch_df)
```

Check the number of NAs in each column

```{r}
colSums(is.na(pitch_df))
```

Lets look and see what the percentage of NAs is per column

```{r}
colMeans(is.na(pitch_df)) * 100  
```

Due to the percentage of missing values being so small in the dataset, we will just remove them all without further investigation

```{r}
na_rows <- apply(is.na(pitch_df), 1, any)
pitch_df <- pitch_df[!na_rows, ]
sum(is.na(pitch_df))
```

EDA

```{r}
summary(pitch_df)
```

Lets look at the categorical features and check their value counts

```{r}
table(pitch_df$pitch_type)
```

-   **AB** = Automatic Ball (e.g., pitch-clock violation, or other “auto ball” call)

-   **CH** = Changeup

-   **CU** = Curveball

-   **EP** = Eephus

-   **FA** = Fastball (general fastball label)

-   **FC** = Cutter

-   **FF** = Four-Seam Fastball

-   **FO** = Forkball

-   **FS** = Split-Finger Fastball (Splitter)

-   **FT** = Two-Seam Fastball

-   **IN** = Intentional Ball (Intentionally thrown outside for a walk)

-   **KC** = Knuckle Curve

-   **KN** = Knuckleball

-   **PO** = Pitch-Out

-   **SC** = Screwball

-   **SI** = Sinker

-   **SL** = Slider

-   **UN** = Unknown (classification software couldn’t assign a type)

```{r}
table(pitch_df$type)
```

```{r}
table(pitch_df$code)
```

## Visualize Outliers

Pitch Speed Distribution by Pitch Type

```{r}
library(ggplot2)

ggplot(pitch_df, aes(x=pitch_type, y= start_speed)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Pitch Speed Distribution by Pitch Type")
```

Distribution of Pitch Types

```{r}
ggplot(pitch_df, aes(x = pitch_type)) +
  geom_bar(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Frequency of Different Pitch Types")
```

Distribution of Pitch Speeds

```{r}
ggplot(pitch_df, aes(x = start_speed)) +
  geom_density(fill = "blue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Density Plot of Pitch Speeds", x = "Pitch Speed (mph)", y = "Density")
```

Outliers in Pitch Speed (start)

```{r}
ggplot(pitch_df, aes(y = start_speed)) +
  geom_boxplot(fill = "lightblue") +
  theme_minimal() +
  labs(title = "Boxplot of Pitch Speeds", y = "Pitch Speed (mph)")
```

Histograms for pitch speed, and spin rate

```{r}
ggplot(pitch_df, aes(x = start_speed)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Pitch Speeds", x = "Pitch Speed (mph)", y = "Count")

ggplot(pitch_df, aes(x = spin_rate)) +
  geom_histogram(binwidth = 50, fill = "darkorange", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Spin Rate", x = "Spin Rate (rpm)", y = "Count")
```

## Correlation Analysis

We will create a correlation matrix among all numeric variables

```{r}
library(corrplot)

numeric_vars <- sapply(pitch_df, is.numeric)
numeric_data <- pitch_df[, numeric_vars]

corr_mat <- cor(numeric_data)

corrplot(
  corr_mat,
  method = "circle",
  type = "upper",
  order = "hclust", #hierachical clustering of vars
  tl.col = "black", #text color
  tl.cex = 0.7, # text size
  tl.srt = 45, #rotation
  col = colorRampPalette(c("navy", "white", "firebrick3"))(200) #Set a palette
)
  
```
